A formal approach but practical approach in train regulation

### 2025-06-25 paper/main.tex
première révision de règle

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The three states of the guard are \guardR, \guardS, and \guardT.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[draw] (trains) {Train movements (\guardT)};
		\node[draw, above right=of trains] (signals1) {Signals update (\guardS)};
		\node[draw, below right=of trains] (regulator) {Regulator reaction (\guardR)};

		\draw[-stealth] (trains) to[bend right=20] (signals1);
		\draw[-stealth] (signals1.west) to[bend right=20] (trains.north);
		\draw[-stealth] (trains.south) to[bend right] (regulator.west);
		\draw[-stealth] (regulator) to[bend right] (signals1);
	\end{tikzpicture}
	\caption{State machine of the guard.}
	\label{fig:state_machine_guard}
\end{figure}

Train events are not dealt with instantly by the regulator (e.g. two train events can happen before the regulator begins to react to the first one). To characterise this behaviour, train events are stored in a buffer until the regulator uses them. In our model, a train event represent a train going from one block to another. When such thing happens, the train sends a message with its identifier to the regulator. A \emph{train buffer} \bufTrain is a sequence of such events, i.e. of train identifiers.
Similarly, the regulator communicates with signals with a \emph{signal buffer} \bufSig, which is a list of tuples $\tuple{\bid{id}, c}$, indicating that both signals of block $\bid{id}$ should be turned to color $c$. Let \bufTrainSet (resp. \bufSigSet) denote the set of train (resp. signal) buffers.

@@

States of the system are triples from $\{\guardR, \guardS, \guardT\} \times \bufTrainSet \times \bufSigSet \times \modelSet$. Our reduction relation is a binary relation on system states, noted $\reduces$: $S_1 \reduces S_2$ means that state $S_1$ can change to $S_2$ (note that this relation is not a function: $S_1$ can reduce to multiple states, due to non-determinism). We define $\reduces$ as the least relation induced by the inference rules presented hereafter.

@@

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\hspace{4mm}  \redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\push{\bufTrain}{\tuple{\tid{id}, \bid{id}}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

The last two rules for trains describe what happens when a train actually moves, i.e. when its position changes. Two cases can occur, depending on the current order of the train: in \su{d}{N}, the $\posFmt{N}$ parameter is a list of blocks the train is about to encounter (including the current block), while going in direction \dirFmt{d}. Two cases occur, depending on whether \posFmt{N} contains future blocks or not. If \posFmt{N} contains only the upcoming block, then upon leaving the block, the order is removed (rule \ruleRef{UntilNext}); otherwise the upcoming block is simply popped from \posFmt{N} (rule \ruleRef{Until}). Notice that in both cases, the state of the guard changes to \guardS, meaning that signals will be updated immediately after that rule.\todo{Introduire/formaliser N} Also, both rules create an event, pushed in the buffer, informing the regulator that the train has changed block.

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}
Notice that, in \ruleRef{UntilNext}, no message is added to the buffer.
\todo{@Lucas, expliquer pourquoi c'est pas nécessaire.}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{StartEvent}] {
		\buflen{B} \neq 0
	}{
		\redTuple{\guardT}{\bufTrain}{\bufSig}{R} \reduces \redTuple{\guardR}{\bufTrain}{\bufSig}{R}
	}
\end{mathpar}

@@

Rule \ruleRef{Turn} triggers when a \turn{\sid{id}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\turn{\sid{id}}{s}}{A^\prime}
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
	\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T, W}}{\updateSignals{\switches}{\sid{id}}{s}}{\signals}}
	}
\end{mathpar}

@@

To each safe step $X = (\vec{x}_i)_{i \in \trains}$, we can associate a system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$ with $\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}$ defined as follows: 
\begin{itemize}
\item for each train $\tid{id}$, $\trainSeq[id] = \trainTuple{\tid{id}}{\bid{id}}{d}{\su{d}{x^{\tid{id}}_{k_{\tid{id}}}}}$ with $\langle d,\sigma\rangle = \text{priv}(\vec{x}^{\tid{id}})$,
\item $R = \langle E,  \bid{id} \in \blocks \mapsto 0, \emptyset\rangle$, $E$ being defined just below,
\item $\sigma$ is the unique network configuration shared by all privileged pairs (here we suppose that there is at least one(!) train),
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$: $\langle \bid{id}, d\rangle \mapsto \siggreen$,
\end{itemize}

Now, let us turn to $E$. Let $\sigma$ be the unique  shared network configuration in pairs and let 
$\vec{y} = \{ \langle x_{k_{\tid{id}}}^\tid{id}, d\rangle \mid \tid{id} \in \trains \text{ and }  \langle d, \sigma\rangle = \text{priv}(\vec{x}^\tid{id}) \}$. 
We define $E(\bot) = (\turn{\sid{id}}{\sigma(\sid{id})})_{\sid{id} \in \turnouts} \cdot (\text{signalred}(x,d))_{(x,d) \in \vec{y}} $.

@@

\paragraph{Experiments}  With TLA+, we checked that any safe step is actually reliable.  On our network, one finds 58 368 safe steps, all being reliable. Checking all the path took XXX seconds on YYY. \todo{XXX= nombre de seconde, YYY = décrire la machine (processeur, RAM, etc)}

@@

Two states $S$ and $S'$ are said to be \emph{compatible} whenever:
\begin{itemize}
\item for any train $\tid{id}$, we have $\mathtt{dst}(S, \tid{id})  =  \mathtt{src}(S', \tid{id})$,
\item for any normal form $S \reduces^! S''$, we have $S''.\sigma = S'.\sigma$.
\end{itemize}
In other words, at the arrival of $S$, trains are at the starting position of $S'$ and so are switches. We define the composition on compatible states. Actually, we describe composition of a system state and a safe step. So, from the remaining of the section, we suppose $S$ and $S'$ be compatible state, $S'$ being a safe step. We aim to compute their composition $S''  = \redTuple{x}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regulator}{\switches}{\signals}}$.

>>>>

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The four states of the guard are \guardI, \guardR, \guardS, and \guardT.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[] (zero) {};
		\node[draw, below=of zero] (initial) {Initial state (\guardI)};
		\node[draw, right=of initial] (trains) {Train movements (\guardT)};
		\node[draw, above right=of trains] (signals1) {Signals update (\guardS)};
		\node[draw, below right=of trains] (regulator) {Regulator reaction (\guardR)};

		\draw[-stealth] (zero) to (initial);
		\draw[-stealth] (initial) to (trains);
		\draw[-stealth] (trains) to[bend right=20] (signals1);
		\draw[-stealth] (signals1.west) to[bend right=20] (trains.north);
		\draw[-stealth] (trains.south) to[bend right] (regulator.west);
		\draw[-stealth] (regulator) to[bend right] (signals1);
	\end{tikzpicture}
	\caption{State machine of the guard.}
	\label{fig:state_machine_guard}
\end{figure}

In our model, a train event represents a train going from one block to another. Train events are not dealt with instantly by the regulator (e.g. two train events can happen before the regulator begins to react to the first one). To characterise this behaviour, train events are stored in a buffer until the regulator uses them. When such thing happens, the train sends a message with its identifier to the regulator. A \emph{train buffer} \bufTrain is a sequence of such events, i.e. of train identifiers.
Similarly, the regulator communicates with signals with a \emph{signal buffer} \bufSig, which is a list of tuples $\tuple{\bid{id}, c}$, indicating that both signals of block $\bid{id}$ should be turned to value $c$. Let \bufTrainSet (resp. \bufSigSet) denote the set of train (resp. signal) buffers.

@@

States of the system are triples from $\{\guardI, \guardR, \guardS, \guardT\} \times \bufTrainSet \times \bufSigSet \times \modelSet$. Our reduction relation is a binary relation on system states, noted $\reduces$: $S_1 \reduces S_2$ means that state $S_1$ can change to $S_2$ (note that this relation is not a function: $S_1$ can reduce to multiple states, due to non-determinism). We define $\reduces$ as the least relation induced by the inference rules presented hereafter.

@@

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\hspace{4mm}  \redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\push{\bufTrain}{\tuple{\tid{id}, \bid{id}}}}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}
\todo{Est ce qu'on met un évenement pour les départs/arrivée? }

The last two rules for trains describe what happens when a train actually moves, i.e. when its position changes. Two cases can occur, depending on the current order of the train: in \su{d}{N}, the $\posFmt{N}$ parameter is a list of blocks the train is about to encounter (including the current block), while going in direction \dirFmt{d}. %Two cases occur, depending on whether \posFmt{N} contains future blocks or not. 
If \posFmt{N} contains only the upcoming block, then upon leaving the block, the order is removed (rule \ruleRef{UntilNext}); otherwise the upcoming block is simply popped from \posFmt{N} (rule \ruleRef{Until}). Notice that in both cases, the state of the guard changes to \guardS, meaning that signals will be updated immediately after that rule.\todo{Introduire/formaliser N} Also, both rules create an event, pushed in the buffer, informing the regulator that the train has changed block.

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}
Notice that, in \ruleRef{UntilNext}, no message is added to the buffer.
\todo{@Lucas, expliquer pourquoi c'est pas nécessaire. \\ @Martin, cela semble contradictoire avec la dernière phrase avant les règles.}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{StartEvent}] {
		\buflen{\bufTrain} \neq  0
	}{
		\redTuple{\guardT}{\bufTrain}{\bufSig}{R} \reduces \redTuple{\guardR}{\bufTrain}{\bufSig}{R}
	}
\end{mathpar}

@@

Rule \ruleRef{Turn} triggers when a \turn{\sid{id}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\handlerOf{E}{\tid{id}} = \langle \bid{id}, \regConcat{\turn{\sid{id}}{s}}{A^\prime}\rangle
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regulator}{\switches}{\signals}}
	\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regulator[E \leftarrow E[ \tid{id} \leftarrow \langle \bid{id}, A'\rangle] ] }{\updateSignals{\switches}{\sid{id}}{s}}{\signals}}
	}
\end{mathpar}

@@

To each safe step $X = (\vec{x}_i)_{i \in \trains}$, just below, we define the system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$. Informally, we set turnouts to their position according to $\sigma$ during the initial step  and we lock signals to red for each train's destination in its intended direction.  
\begin{itemize}
\item for each train $\tid{id}$, $\trainSeq[id] = \trainTuple{\tid{id}}{\bid{id}}{d}{\su{d}{x^{\tid{id}}_{k_{\tid{id}}}}}$ with $\langle d,\sigma\rangle = \text{priv}(\vec{x}^{\tid{id}})$,
\item $R = \langle E,  \bid{id} \in \blocks \mapsto 0, \emptyset\rangle$, $E$ being defined just below,
\item $\sigma$ is the unique network configuration shared by all privileged pairs (here we suppose that there is at least one(!) train),
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$: $\langle \bid{id}, d\rangle \mapsto \siggreen$,
\end{itemize}

Now, let us turn to $E$. Let $\sigma$ be the unique  shared network configuration in privileged pairs and let 
$\vec{y} = \{ \langle x_{k_{\tid{id}}}^\tid{id}, d\rangle \mid \tid{id} \in \trains \text{ and }  \langle d, \sigma\rangle = \text{priv}(\vec{x}^\tid{id}) \}$. 
We define $E(\bot) = (\turn{\sid{id}}{\sigma(\sid{id})})_{\sid{id} \in \turnouts} \cdot (\text{signalred}(x,d))_{(x,d) \in \vec{y}} $.

@@

\paragraph{Experiments}  With TLA+, we checked that any safe step is actually (not surprisingly!) reliable.  On our network, one finds 58 368 safe steps, all being reliable. Checking all the path took XXX seconds on YYY. \todo{XXX= nombre de seconde, YYY = décrire la machine (processeur, RAM, etc)}

@@

Two states $S$ and $S'$ are said to be \emph{compatible} whenever  for any train $\tid{id}$, we have $\mathtt{dst}(S, \tid{id})  =  \mathtt{src}(S', \tid{id})$. In other words, train must start in $S'$ at destination in $S$. We define the composition only compatible states. Actually, since we are interested by safe steps composition, we restrict our attention to their compositions.

we describe composition of a system state and a safe step. So, from the remaining of the section, we suppose $S$ and $S'$ be compatible state, $S'$ being a safe step. We aim to compute their composition $S''  = \redTuple{x}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regulator}{\switches}{\signals}}$.
